---
title: "Functional PCA"
output:
  html_document: 
    code_folding: hide
    toc: true
    toc_float: true
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 8,
  fig.height = 4
)
```


This page contains a combination of traditional lecture materials (slides) and code demonstrating the relevant methods. The short course will proceed by working through both. We will use several recent packages in our examples; see the [About](./About.html) page for information about the package versions.

```{r setup}
library(tidyverse)
library(grpreg)
library(splines)
library(refund)
library(refund.shiny)
library(reshape2)
library(plotly)
```

In this section, we will use the [NHANES_JSM_2019](./shortcourse_data.html) data to illustrate scalar on function regression (sofr).

## NHANES data

The participant-specific daily activity profiles have been created by i) log-transforming minute level counts per minute, ii) calculating hourly total activity count (hTAC), iii) averaging the hTAC over all available days. Thus, a participant-specific daily activity profile is represented by 24 points.

First, we to plot all participant-specific daily activity profiles:

$$x_i(t), t = 1,\ldots, 24$$ 


```{r nhanes_all_subjects, eval = TRUE}
#load("./DataCode/JSM_2019_NHANES.RDA")
load("./DataCode/NHANES_JSM_2019.RData")

hTAC = as.matrix(hTAC)
n = dim(hTAC)[1] # number of participants
col.light = rgb(0, 0, 0, alpha = 5, maxColorValue = 255)
x.range = 1:24

## plot all profiles 
plot(x.range, hTAC[1,], type = "n", ylim = range(hTAC), xaxt = "n", ylab = "Log of Activity Counts", xlab = "Hour", col = col.light, main = "Diurnal Profiles")
for(i in 2:n) lines(1:24, hTAC[i,], col = col.light, type = "l", lwd = 1)
axis(1, at = c(1, 4, 8, 12, 16, 20, 24), labels = c("01:00", "04:00", "08:00", "12:00", "16:00", "20:00", "24:00"))
abline(v = 1:24, col = "gray", lty = "dotted")
```

## Marginal information

One way to look at the marginal information is by looking at **marginal hour-specific means and standard deviations**. Below, we add the mean curve and plus/minus 1.96 times standard deviations curves

```{r nhanes_all_subjects_01, eval = TRUE}
hTAC.mean <- apply(hTAC, 2, mean)
hTAC.sd <- apply(hTAC, 2, sd)

## plot all profiles 
plot(x.range, hTAC[1,], type = "n", ylim = range(hTAC), xaxt = "n", ylab = "Log of Activity Counts", xlab = "Hour", col = col.light, main = "Diurnal Profiles")
for(i in 2:n) lines(x.range, hTAC[i,], col = col.light, type = "l", lwd = 1)
lines(x.range, hTAC.mean, col = "blue", type = "l", lwd = 2)
lines(x.range, hTAC.mean + 1.96*hTAC.sd, col = "blue", type = "l", lty = "dashed", lwd = 2)
lines(x.range, hTAC.mean - 1.96*hTAC.sd, col = "blue", type = "l", lty = "dashed", lwd = 2)
axis(1, at = c(1, 4, 8, 12, 16, 20, 24), labels = c("01:00", "04:00", "08:00", "12:00", "16:00", "20:00", "24:00"))
abline(v = 1:24, col = "gray", lty = "dotted")
```

### Sample summaries

We can calculate **hour-specific sample mean** as 

  $$\bar{x}(t)= \frac{1}{n}\sum_{i=1}^n x_i(t)$$

the **hour-specific sample variance** as 

  $$\widehat{Var}(x(t)) = \frac{1}{n}\sum_{i=1}^n (x_i(t)-\bar{x}(t))^2$$
  
and the **total sample variation** as 

  $$\sum_{t=1}^T \widehat{Var}(x(t)) = \frac{1}{n}\sum_{t=1}^T 
  \sum_{i=1}^n (x_i(t)-\bar{x}(t))^2$$
  
> PCA treats the total variation as total information

### Sample covariance and correlation matrixes  

**Sample covariance matrix** is defined as


$$\hat{\Sigma}(t,s) = \widehat{Cov}(x(t), x(s)) = \frac{1}{n}\sum_{i=1}^n (x_i(t)-\bar{x}(t))(x_i(s)-\bar{x}(s)), \quad s,t = 1,\ldots, 24$$

```{r nhanes_all_subjects_02, eval = TRUE}
  ## plot the covariance surface
  hTAC.cov = cov(hTAC)
  plot_ly(z = ~hTAC.cov) %>% add_surface()

```
**Sample correlation matrix** is defined as 
  
$$\widehat{Corr}(x(t), x(s)) = \frac{\widehat{Cov}(x(t), x(s))}{\sqrt{\widehat{Cov}(x(t))\widehat{Cov}(x(s))}}, \quad s,t = 1,\ldots, 24$$
  
  
```{r nhanes_all_subjects_03, eval = TRUE}
  ## plot the covariance surface
  hTAC.cor = cor(hTAC)
  plot_ly(z = ~hTAC.cor) %>% add_surface()
```  


## Intuition behind PCA

### De-meaning


```{r nhanes_all_subjects_04, eval = TRUE}
# create de-meaned diurnal profiles
hTAC.demean = hTAC - matrix(apply(hTAC, 2, mean), n, 24, byrow = TRUE)

## plot all profiles 
plot(x.range, hTAC.demean[1,], type = "n", ylim = c(-300,300), xaxt = "n", ylab = "Log of Activity Counts", xlab = "Hour", col = col.light, main = "Diurnal Profiles")
for(i in 2:n) lines(1:24, hTAC.demean[i,], col = col.light, type = "l", lwd = 1)
axis(1, at = c(1, 4, 8, 12, 16, 20, 24), labels = c("01:00", "04:00", "08:00", "12:00", "16:00", "20:00", "24:00"))
abline(h = 0, col = "blue", lwd = 3)
lines(1.96*hTAC.sd, col = "blue", lwd = 3, type = "l", lty = "dashed")
lines(-1.96*hTAC.sd, col = "blue", lwd = 3, type = "l", lty = "dashed")
abline(v = 1:24, col = "gray", lty = "dotted")
```

### Principal components

Find "principal" patterns that explain largest proportions of **total variation (= information)**. Principal components = "principal" patterns (PC1 = red, PC2 = green, PC3 = blue) (1-2-3 = rgb)

```{r nhanes_all_subjects_07, eval = TRUE}
s = eigen(hTAC.cov, symmetric = T)
phi = s$vectors
col = c("red", "green", "blue")
plot(1:24, phi[,1], ylim = c(-0.4, 0.4), xaxt = "n", xlab = "Time of Day", ylab = "AU", lwd = 3, col = col[1], type = "l")
for (k in 2:3) lines(1:24, phi[,k], lwd = 3, col = col[k], type = "l")
axis(1, at = c(1, 4, 8, 12, 16, 20, 24), labels = c("01:00", "04:00", "08:00", "12:00", "16:00", "20:00", "24:00"))
abline(v = 1:24, col = "gray", lty = "dotted")
abline(h = 0, col = "black", lty = "dotted", lwd = 2)
```

### PC1 

PC1 explains 37.5% of total variation (note the local behaviour)

```{r nhanes_all_subjects_08, eval = TRUE}

# create de-meaned diurnal profiles, calculate 1st PC and subtract its contribution
hTAC.demean.cov = cov(hTAC.demean)
s = eigen(hTAC.demean.cov, symmetric = T)
phi = s$vectors
PC1 = phi[,1]
PC1.score = as.matrix(hTAC.demean) %*% phi[,1]
hTAC.PC1 <- PC1.score %*% phi[,1]
hTAC.demean.minus.PC1 = hTAC.demean - hTAC.PC1


## plot all profiles 
plot(x.range, hTAC.PC1[1,], type = "n", ylim = c(-300,300), xaxt = "n", ylab = "Log of Activity Counts", xlab = "Hour", col = col.light, main = "Diurnal Profiles")
for(i in 2:n) lines(1:24, hTAC.PC1[i,], col = col.light, type = "l", lwd = 1)
axis(1, at = c(1, 4, 8, 12, 16, 20, 24), labels = c("01:00", "04:00", "08:00", "12:00", "16:00", "20:00", "24:00"))

abline(h = 0, col = "blue", lwd = 3)
lines(1.96*hTAC.sd, col = "blue", lwd = 3, type = "l", lty = "dashed")
lines(-1.96*hTAC.sd, col = "blue", lwd = 3, type = "l", lty = "dashed")
abline(v = 1:24, col = "gray", lty = "dotted")
```

### PC2

PC2 explains 22.5% of total variation (but note the local behaviour)

```{r nhanes_all_subjects_09, eval = TRUE}
# create de-meaned diurnal profiles, calculate 1st PC and subtract its contribution
hTAC.demean.cov = cov(hTAC.demean)
s = eigen(hTAC.demean.cov, symmetric = T)
phi = s$vectors
PC2 = phi[,2]
PC2.score = as.matrix(hTAC.demean) %*% PC2
hTAC.PC2 <- PC2.score %*% PC2
hTAC.demean.minus.PC2 = hTAC.demean - hTAC.PC2


## plot all profiles 
plot(x.range, hTAC.PC2[1,], type = "n", ylim = c(-300,300), xaxt = "n", ylab = "Log of Activity Counts", xlab = "Hour", col = col.light, main = "Diurnal Profiles")
for(i in 2:n) lines(1:24, hTAC.PC2[i,], col = col.light, type = "l", lwd = 1)
axis(1, at = c(1, 4, 8, 12, 16, 20, 24), labels = c("01:00", "04:00", "08:00", "12:00", "16:00", "20:00", "24:00"))
abline(h = 0, col = "blue", lwd = 3)
lines(1.96*hTAC.sd, col = "blue", lwd = 3, type = "l", lty = "dashed")
lines(-1.96*hTAC.sd, col = "blue", lwd = 3, type = "l", lty = "dashed")
abline(v = 1:24, col = "gray", lty = "dotted")
```

### PC3

PC3 explains 10.7% of total variation (but note the local behaviour)

```{r nhanes_all_subjects_10, eval = TRUE}
# create de-meaned diurnal profiles, calculate 1st PC and subtract its contribution
PC3 = phi[,3]
PC3.score = as.matrix(hTAC.demean) %*% PC3
hTAC.PC3 <- PC3.score %*% PC3
hTAC.demean.minus.PC3 = hTAC.demean - hTAC.PC3

## plot all profiles 
plot(x.range, hTAC.PC3[1,], type = "n", ylim = c(-300,300), xaxt = "n", ylab = "Log of Activity Counts", xlab = "Hour", col = col.light, main = "Diurnal Profiles")
for(i in 2:n) lines(1:24, hTAC.PC3[i,], col = col.light, type = "l", lwd = 1)
axis(1, at = c(1, 4, 8, 12, 16, 20, 24), labels = c("01:00", "04:00", "08:00", "12:00", "16:00", "20:00", "24:00"))
abline(h = 0, col = "blue", lwd = 3)
lines(1.96*hTAC.sd, col = "blue", lwd = 3, type = "l", lty = "dashed")
lines(-1.96*hTAC.sd, col = "blue", lwd = 3, type = "l", lty = "dashed")
abline(v = 1:24, col = "gray", lty = "dotted")
```

Let us consider a specific subject and plot subject's original profile, the profile representation via PC1, PC2, and PC3.

```{r nhanes_all_subjects_07_a, eval = TRUE}
s.id = 1
plot(1:24, hTAC[s.id,], ylim = c(-300,300), xaxt = "n", xlab = "Time of Day", ylab = "Log of AC", lwd = 3, col = "black", type = "l")
lines(1:24, hTAC.mean, col = "gray", lwd = 3, type = "l")
lines(1:24, hTAC.PC1[s.id,], col = col[1], lwd = 3, type = "l")
lines(1:24, hTAC.PC2[s.id,], col = col[2], lwd = 3, type = "l")
lines(1:24, hTAC.PC3[s.id,], col = col[3], lwd = 3, type = "l")
axis(1, at = c(1, 4, 8, 12, 16, 20, 24), labels = c("01:00", "04:00", "08:00", "12:00", "16:00", "20:00", "24:00"))
abline(v = 1:24, col = "gray", lty = "dotted")
abline(h = 0, col = "black", lty = "dotted", lwd = 2)

## corresponding scores for this subject
PC1.score[s.id]
PC2.score[s.id]
PC3.score[s.id]
```
We also plot the original profile and demeaned profile, and profile after subtracting PC1, PC2, PC3 contributions.

```{r nhanes_all_subjects_07_b, eval = TRUE}
s.id = 1
plot(1:24, hTAC[s.id,], ylim = c(-300,300), xaxt = "n", xlab = "Time of Day", ylab = "Log of AC", lwd = 3, col = "black", type = "l")
lines(1:24, hTAC.demean[s.id,], col = "gray", type = "l")
lines(1:24, hTAC.demean.minus.PC1[s.id,], col = col[1], lwd = 3, type = "l")
lines(1:24, hTAC.demean.minus.PC2[s.id,], col = col[2], lwd = 3, type = "l")
lines(1:24, hTAC.demean.minus.PC3[s.id,], col = col[3], lwd = 3, type = "l")
axis(1, at = c(1, 4, 8, 12, 16, 20, 24), labels = c("01:00", "04:00", "08:00", "12:00", "16:00", "20:00", "24:00"))
abline(v = 1:24, col = "gray", lty = "dotted")
abline(h = 0, col = "black", lty = "dotted", lwd = 2)
```

```{r nhanes_all_subjects_11, eval = TRUE}
plot(x.range, hTAC.demean[1,], type = "n", ylim = c(-300,300), xaxt = "n", ylab = "Log of Activity Counts", xlab = "Hour", col = col.light, main = "Diurnal Profiles")
for(i in 2:n) lines(1:24, hTAC.demean[i,], col = col.light, type = "l", lwd = 1)
axis(1, at = c(1, 4, 8, 12, 16, 20, 24), labels = c("01:00", "04:00", "08:00", "12:00", "16:00", "20:00", "24:00"))
abline(h = 0, col = "blue", lwd = 3)
lines(1.96*hTAC.sd, col = "blue", lwd = 3, type = "l", lty = "dashed")
lines(-1.96*hTAC.sd, col = "blue", lwd = 3, type = "l", lty = "dashed")
abline(v = 1:24, col = "gray", lty = "dotted")
```

Subtract PC1 that explains 37.5% of total variation

```{r nhanes_all_subjects_12, eval = TRUE}
PC1.score = as.matrix(hTAC.demean) %*% PC1
hTAC.PC1 <- PC1.score %*% PC1
hTAC.demean.minus.PC1 = hTAC.demean - hTAC.PC1


## plot all profiles 
plot(x.range, hTAC.demean.minus.PC1[1,], type = "n", ylim = c(-300,300), xaxt = "n", ylab = "Log of Activity Counts", xlab = "Hour", col = col.light, main = "Diurnal Profiles")
for(i in 2:n) lines(1:24, hTAC.demean.minus.PC1[i,], col = col.light, type = "l", lwd = 1)
axis(1, at = c(1, 4, 8, 12, 16, 20, 24), labels = c("01:00", "04:00", "08:00", "12:00", "16:00", "20:00", "24:00"))
abline(h = 0, col = "blue", lwd = 3)
lines(1.96*hTAC.sd, col = "blue", lwd = 3, type = "l", lty = "dashed")
lines(-1.96*hTAC.sd, col = "blue", lwd = 3, type = "l", lty = "dashed")
abline(v = 1:24, col = "gray", lty = "dotted")
```

Subtract PC1 and PC2 that explain 60% of total variation

```{r nhanes_all_subjects_13, eval = TRUE}
s = eigen(hTAC.demean.cov, symmetric = T)
phi = s$vectors

## PC1
PC1 = phi[,1]
PC1.score = as.matrix(hTAC.demean) %*% PC1
hTAC.PC1 <- PC1.score %*% PC1

## PC2
PC2 = phi[,2]
PC2.score = as.matrix(hTAC.demean) %*% PC2
hTAC.PC2 <- PC2.score %*% PC2
hTAC.demean.minus.PC12 = hTAC.demean - hTAC.PC1 - hTAC.PC2

## plot all profiles 
plot(x.range, hTAC.demean.minus.PC12[1,], type = "n", ylim = c(-300,300), xaxt = "n", ylab = "Log of Activity Counts", xlab = "Hour", col = col.light, main = "Diurnal Profiles")
for(i in 2:n) lines(1:24, hTAC.demean.minus.PC12[i,], col = col.light, type = "l", lwd = 1)
axis(1, at = c(1, 4, 8, 12, 16, 20, 24), labels = c("01:00", "04:00", "08:00", "12:00", "16:00", "20:00", "24:00"))
abline(h = 0, col = "blue", lwd = 3)
lines(1.96*hTAC.sd, col = "blue", lwd = 3, type = "l", lty = "dashed")
lines(-1.96*hTAC.sd, col = "blue", lwd = 3, type = "l", lty = "dashed")
abline(v = 1:24, col = "gray", lty = "dotted")
```
Subtract PC1, PC2, and PC3 that explain 70.8% of total variation

```{r nhanes_all_subjects_14, eval = TRUE}
hTAC.demean.cov = cov(hTAC.demean)
s = eigen(hTAC.demean.cov, symmetric = T)
phi = s$vectors

## PC1
PC1 = phi[,1]
PC1.score = as.matrix(hTAC.demean) %*% PC1
hTAC.PC1 <- PC1.score %*% PC1

## PC2
PC2 = phi[,2]
PC2.score = as.matrix(hTAC.demean) %*% PC2
hTAC.PC2 <- PC2.score %*% PC2

## PC3
PC3 = phi[,3]
PC3.score = as.matrix(hTAC.demean) %*% PC3
hTAC.PC3 <- PC3.score %*% PC3
hTAC.demean.minus.PC123 = hTAC.demean - hTAC.PC1 - hTAC.PC2 - hTAC.PC3

## plot all profiles 
plot(x.range, hTAC.demean.minus.PC123[1,], type = "n", ylim = c(-300,300), xaxt = "n", ylab = "Log of Activity Counts", xlab = "Hour", col = col.light, main = "Diurnal Profiles")
for(i in 2:n) lines(1:24, hTAC.demean.minus.PC123[i,], col = col.light, type = "l", lwd = 1)
axis(1, at = c(1, 4, 8, 12, 16, 20, 24), labels = c("01:00", "04:00", "08:00", "12:00", "16:00", "20:00", "24:00"))
abline(h = 0, col = "blue", lwd = 3)
lines(1.96*hTAC.sd, col = "blue", lwd = 3, type = "l", lty = "dashed")
lines(-1.96*hTAC.sd, col = "blue", lwd = 3, type = "l", lty = "dashed")
abline(v = 1:24, col = "gray", lty = "dotted")
```
### **Scree-plot**

```{r nhanes_all_subjects_05, eval = TRUE}
s = eigen(hTAC.cov, symmetric = T)
lambda = s$values
plot(1:24, 100*lambda/sum(lambda), xlab = "PC", ylab = "% of total variation explained", lwd = 4, col = "blue")
abline(v = 1:24, col = "gray", lty = "dotted")
abline(h = (1:10)*10, col = "gray", lty = "dotted")
```


```{r nhanes_all_subjects_06, eval = TRUE}
plot(1:24, 100*cumsum(lambda)/sum(lambda), xlab = "PC", ylab = "cumulative % of total variation explained",lwd = 4, col = "blue")
abline(v = 1:24, col = "gray", lty = "dotted")
abline(h = (1:10)*10, col = "gray", lty = "dotted")
```



## Functional PCA 

Classical PCA does not use smoothness. Functional PCA smooths the sample covariance matrix and then finds corresponding PCs. R package "refund" performs functional PCA.

```{r nhanes_all_subjects_15, eval = TRUE}
s = fpca.sc(hTAC, var = TRUE, npc = 6)
x.mean = s$mu
phi.fpca = s$efunctions
lambda.fpca = s$evalues
sigma.fpca = s$sigma2
```

Let us compare PC1 (blue dots) and fPC1 (red curve)

```{r nhanes_all_subjects_16, eval = TRUE}
# create de-meaned diurnal profiles, calculate 1st PC and subtract its contribution
s.svd = eigen(hTAC.demean.cov, symmetric = T)
phi.svd = s.svd$vectors

## plotting smooth vs non-smooth PCs
plot(x.range, phi.svd[,1],  ylim = c(-0.4, 0.4), xaxt = "n", ylab = "AU", xlab = "Hour", col = "blue", main = "Compare PCs vs fPCs")
lines(1:24, phi.fpca[,1]/sqrt(sum(phi.fpca[,1]^2)), col = "red", type = "l", lwd = 1, lty = "dashed")
axis(1, at = c(1, 4, 8, 12, 16, 20, 24), labels = c("01:00", "04:00", "08:00", "12:00", "16:00", "20:00", "24:00"))
abline(h = 0, col = "gray", lwd = 2)
abline(v = 1:24, col = "gray", lty = "dotted")
```

Let us compare PC2 (blue dots) vs fPC2 (red curve)



```{r nhanes_all_subjects_17, eval = TRUE}
# create de-meaned diurnal profiles, calculate 1st PC and subtract its contribution
s.svd = eigen(hTAC.demean.cov, symmetric = T)
phi.svd = s.svd$vectors

## plotting smooth vs non-smooth PCs
plot(x.range, phi.svd[,2],  ylim = c(-0.4, 0.4), xaxt = "n", ylab = "AU", xlab = "Hour", col = "blue", main = "Compare PCs vs fPCs")
lines(1:24, phi.fpca[,2]/sqrt(sum(phi.fpca[,2]^2)), col = "red", type = "l", lwd = 1, lty = "dashed")
axis(1, at = c(1, 4, 8, 12, 16, 20, 24), labels = c("01:00", "04:00", "08:00", "12:00", "16:00", "20:00", "24:00"))
abline(h = 0, col = "gray", lwd = 2)
abline(v = 1:24, col = "gray", lty = "dotted")
```

> Main reasons for not seeing larger differences: pre-smoothing (binning in hour epochs), smaller samples likely results in larger deviations.

### Technical details 

fPCA wants to find a sequence of orthonormal functions $\phi_1(t),\ldots, \phi_K(t)$ that sequentially explains most of **total variation = total information**
  
  **Step 1:** Find $\phi_1(t)$ that maximizes $\sum_{i=1}^n \xi^2_{i1}$, where

scores/projections $\xi_{i1} = <x_i, \phi_1> = \int_1^{24} x_i(t)\phi_1(t) dt \quad \mbox{and}\quad \int_1^{24} \phi^2_1(t) dt = 1$ keep the information.
  
  **Step k:** Find $\phi_k(t)$ that maximizes $\sum_{i=1}^n \xi^2_{ik}$
  and  $<\phi_k,\phi_s> = 0, s < k$
  
 scores $\xi_{ik} = <x_i, \phi_k> = \int_1^{24} x_i(t)\phi_k(t) dt\quad \mbox{and} \quad \int_1^{24} \phi^2_k(t) dt = 1$
  
Thus, each observed diurnal pattern is substituted with its fPCA approximation $$x_i(t) \approx \mu(t) + \phi_1(t)\xi_{i1}+\phi_2(t)\xi_{i2}+\phi_3(t)\xi_{i3} $$
 $$\phi_1(t),\phi_2(t), \phi_3(t)$$ are **principal functions** (data-driven population-level basis)
+ $\xi_{i1},\xi_{i2},\xi_{i3}$ are **principal scores** (subject-specific coordinates)

### Dimension-reduction via fPCA

Instead of dealing with **24 correlated numbers** $(x_i(1), x_i(2),\ldots, x_i(24))$, we can now focus only on **3 uncorrelated scores** $\xi_{i1}, \xi_{i2}, \xi_{i3}$ that are ordered in terms of the amount of preserved information and combined preserve roughly 70% of total variation.